Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-11-06 manuscript/manuscript.tex
Top to bottom review. Mostly cleanup, spelling, grammar, etc.
Still need to work on the sections at the end. Part 4.

\hyphenpenalty=750

@@

This paper examines the challenges of open data curation through a
case study of the NYC 311 Service Request (SR) dataset, addressing issues 
of data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government-released open data. These principles 
aim to enhance data management practices and ensure 
the ongoing utility of open data. The paper concludes with 
actionable recommendations for enhancing data curation and outlines
general principles for the effective release of open data.

@@

The City of New York (NYC) has emerged as a leader in the open data movement,
marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2,700 datasets
from 80 different city agencies. This resource has become invaluable
for researchers across various fields and has significantly enhanced
local government transparency. Popular datasets include information on
restaurant health inspection violations, car crashes, high school and
college enrollment statistics, jail inmate charges, and the location
of city-wide free Internet access points. These datasets have been
applied in civil life in various ways, such as mapping car crashes
involving pedestrians and visualizing high school and college
enrollment trends. Furthermore, they have enabled significant research
across multiple domains, including health \citep{cantor2018facets,
  shankar2021data}, urban development \citep{neves2020impacts}, and
transportation \citep{gerte2019understanding}, aiding in the
understanding and addressing of complex urban challenges.

Data curation, the process of organizing, maintaining, and ensuring
the quality of datasets, plays a crucial role in maximizing the
utility of open data. Proper curation ensures that datasets remain
consistent, accurate, and useful for diverse applications. For example, 
well-curated data is essential for machine learning systems, which 
require high-quality data to produce reliable insights 
\citep{polyzotis2019data, jain2020overview}. A critical component of 
data curation is data cleaning, which involves identifying and rectifying 
inconsistencies, errors, and inaccuracies in datasets. While open data 
initiatives have made vast amounts of data available, ensuring the 
reliability of such data hinges on rigorous data cleaning processes. 
Efficient usage of software tools can significantly streamline this 
aspect of curation \citep[e.g.,][]{cody2017cody, van2018statistical}. Poor 
curation may result in issues such as missing data, formatting errors, 
or inconsistent values, leading to biased or inaccurate outcomes 
\citep{geiger2020garbage}. This is particularly critical in domains 
where machine learning is applied to sensitive tasks, such as public 
health or policy \citep{rahm2000data}.

Research into data curation has explored these challenges in\mbox{-}depth. 
Among the earliest discussions, \citet{witt2009constructing} focused 
on developing data curation profiles tailored to specific contexts, 
setting a precedent for targeted data management strategies. 
Addressing broader challenges in data sharing and management, 
\citet{borgman2012conundrum} highlighted the complexities of 
research data distribution, emphasizing the need for robust 
strategies. This is complemented by \citet{hart2016ten}, who outlined 
essential principles for effective data management, particularly 
emphasizing meticulous curation practices. The utility of curated 
open data is vividly illustrated in public health and global challenges, 
where \citet{cantor2018facets} demonstrated the utility of curated 
data in evaluating community health determinants, and 
\citet{shankar2021data} observed its critical role during the 
COVID-19 pandemic in managing collective responses.

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request (SR) Data as a case study. This dataset serves as a prime 
example for examining key issues in data curation, including data 
validity, consistency, and curation efficiency. 
We illustrate these points with live examples drawn from our 
processing of the 311 SR data. Secondly, building upon insights 
gained from this case study, we propose a set of data curation 
principles tailored for government-released open data. These 
principles are designed to address the unique challenges 
and requirements observed in the curation of such datasets.

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over a 2-year period from 2022 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
 investigates field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the Data Dictionary. Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system (the biggest expansion in their history), and a 
record 3.23 million service requests in 2023.

Today, the NYC 311 data system manages over 3 million service
requests (SRs) per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, provides technological support for the
open data system and data portal. Input data is sourced from 17 different
NYC Agencies.

\begin{comment}
Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usage for numerous
third-party users. These challenges are addressed by various agency
open data managers and OTI.
\end{comment}

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping city governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID-19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies

Our investigation utilizes an SR dataset covering  2-years (2022-2023).
This dataset is approximately 3.7 GB in size and contains over 
6.4 million rows with each row representing one SR. It contains 
41 columns of data. It is used to conduct detailed analysis 
in areas such as data consistency, data validity, redundant fields, 
etc.. Instructions on downloading this file from the 
NYC Open Data Portal are available in the supplemental material.

During the 10-year period from 2014 to 2023,  311 experienced
a 50\% in Service Requests, , reflecting a growing public reliance 
on the system. (As a note, during this 10-year timeframe, the 
population of NYC grew just 0.6\%.). The most significant increase
in activity occurred in 2020, when the number of requests 
surged to 3.23 million during the COVID-19 pandemic, 
illustrating the essential role of the 311 system in supporting 
NYC residents during times of crisis. The steady 
rise in SRs can be at least partially attributed to the increased 
accessibility of the 311 system via online and mobile 
platforms, as well as heightened public awareness of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in SR volume in future years,  
highlighting the system's evolving role in managing 
both routine city operations and extraordinary events.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
agency, showing the cumulative percentage of SRs handled by each
agency over a the 2-year timeframe. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

\begin{itemize}
    \item New York Police Department (NYPD) - 43\%
    \item Housing Preservation and Development (HPD) - 21\%
    \item New York City Department of Sanitation (DSNY) - 10\%
    \item Department of Transportation (DOT) - 7\%
    \item Department of Environmental Protection (DEP) - 5\%
    \item Department of Parks and Recreation (DPR) - 4\%
\end{itemize}

Collectively these six City agencies handle over 90\% of the total SR 
volume, indicative of the critical role these agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The NYPD alone accounts for a
substantial portion of the total SRs, underscoring its leading role in
addressing complaints related to public safety. The remaining
10\% of SRs is distributed across 10 other
agencies. This concentration of SRs to the ``big six'' agencies 
underscores the necessity for optimal data handling and curation processes 
within these high-volume agencies in order to ensure 
efficiency and responsiveness.

@@

@@

Figure~\ref{fig:SR_complaints}
illustrates how the top 20 complaint types, accounting for 70\% 
of all SRs. Note that the distribution of complaints is skewed, 
with a small number of issues dominating total SR volume. 
Noise-related complaints, including residential, commercial, and 
street noise, when combined make up 22\% of all 
complaints; the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
reveals that after the top 20 complaint types, the remaining 
complaint types are spread thinly across the remaining 30\% of SRs,
suggesting that improving responses to the most frequent 
complaints could have an outsized impact on overall service 
efficiency and resident satisfaction. This chart also provides 
insights into the operational pressures faced by City agencies 
responsible for handling these high-volume complaints.

\section{NYC 311 SR Data Cleansing Issues} 
\label{sec:issues}

Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without thorough data cleansing, the utility
of open data can be limited and perhaps untrustworthy, affecting 
its reliability for research, policy-making, and innovation. The 
main purpose of cleansing open data is to ensure that it is 
accurate, consistent, and usable across multiple platforms and by 
various stakeholders with a myriad of purpose. Data cleansing improves
the trustworthiness of the data and enables more accurate analysis, 
better decision-making,and the integration of data into machine
learning models or other systems.

Many quality criteria are employed to ensure high-quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats; typically 
outlined in a Data Dictionary. An essential aspect of validation 
is domain compliance, where fields must adhere to a predefined set of 
values, such as statuses, state names, or zip codes. Structural 
errors also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary or the 
inconsistent use of blanks, spaces, NA, N/A, or ``<NA>'' to indicate 
missing data. Furthermore, redundant or irrelevant fields can clutter 
datasets, reducing efficiency. Logical inconsistencies are another 
important consideration, such as related fields that violate expected 
relationships, such as a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure reliable and meaningful data.

Here we identify the presence of issues in the NYC 311 SR open data;
not attempting to solve them but rather to highlight their scope and
magnitued. A solution effort should be undertaken only after further 
investigation as to the why and how such issues came about, 
and a discussion as to whether or not it is an actual \textit{error} or 
holds some other status. Coordination with the originating Agency
is paramount.

\subsection{Structural Issues}
\label{sec:structural}

Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can make 
it difficult to analyze the data effectively. Here are some 
characteristics of the 311 SR data set:

\begin{itemize}
	\item There are 41 columns of data for each row, exportable as a CSV file.
	Each row represents one Service Request.

@@

	\item Seven street fields; two pair of which appear to be duplicates

@@

	\item A free-form text field, \texttt{resolution\_description}, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' The additional fields include:

@@

While one can likely infer the meaning and 
derivation of these computed fields, uncertainty remains as to
their derivation and usage. These six fields are not included in
the Data Dictionary, which we have asked to be updated. Our 
analysis shows that these six \texttt{computed} fields have 
significant data validity issues, leading us to treat these six fields 
as experimental and not for suitable for normal use. Accordingly, 
we will not address these fields in our analysis.

\subsection{Missing Data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} field is blank. When counting fields for 
blank or N/A values, they appear to divide into three groups:

@@

The Mostly Empty category includes such fields as
\texttt{taxi\_company\_borough}, \texttt{due\_date},
\texttt{road\_ramp}, and \texttt{bridge\_highway\_name}.
The Partially Empty includes such fields as
\texttt{location\_type}, \texttt{landmark}, 
and \texttt{cross\_street\_1 \& \_2}. And the Few/None Empty includes
\texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. In some analysis efforts, it may be
prudent to inquire as to why some fields are mostly or almost always blank.
Figure~\ref{fig:blank_fields} presents a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

@@

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. Below are the results of 
selected field validations. The latitude and longitude fields were 
found to fall within the geographic boundaries of New York City, and 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the Data Dictionary specifies very few domains of 
acceptable values. Nonetheless, the following fields were tested and found 
to comply with their expected domains, as determined by public usage 
and historical datasets:

@@

But some fields proved problematic as to their 
compliance with a domain of allowable values.

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes ( \texttt{incident\_zip}) 
should validate against the USPS database, which contains 
44,173 valid zip codes. The \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by Agency, it mirrors the \textit{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\paragraph{Created and Closed Date}
\label{sec:negativeduration}
For the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields to be automatically populated by the  
application software when setting an SR status to ``new'' or ``closed.'' 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the \texttt{created\_date} and \texttt{closed\_date} fields that
an automated process would normally exclude, including:

\begin{itemize}
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} creating a ``negative duration''.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. ``1900-01-01''.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match \textbf{to the second} indicating a nonsensical ``zero duration''.
    \item An unusually large number of SRs closed and/or created exactly at midnight 
    and noon, to the second.
\end{itemize}

>>>>

\hyphenpenalty=900

@@

This paper examines the challenges of open data curation through a
case study of the NYC 311 Service Request (SR) dataset, addressing issues 
of data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government\mbox{-}released open data. These principles 
aim to enhance data management practices and ensure 
the ongoing utility of open data. The paper concludes with 
actionable recommendations for enhancing data curation and outlines
general principles for the effective release of open data.

@@

The City of New York (NYC) has emerged as a leader in the open data movement,
marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2,700 datasets
from 80 different city agencies. This resource has become invaluable
for researchers across various fields and has significantly enhanced
local government transparency. Popular datasets include information on
restaurant health inspection violations, car crashes, high school and
college enrollment statistics, jail inmate charges, and the location
of city\mbox{-}wide free Internet access points. These datasets have been
applied in civil life in various ways, such as mapping car crashes
involving pedestrians and visualizing high school and college
enrollment trends. Furthermore, they have enabled significant research
across multiple domains, including health \citep{cantor2018facets,
  shankar2021data}, urban development \citep{neves2020impacts}, and
transportation \citep{gerte2019understanding}, aiding in the
understanding and addressing of complex urban challenges.

Data curation, the process of organizing, maintaining, and ensuring
the quality of datasets, plays a crucial role in maximizing the
utility of open data. Proper curation ensures that datasets remain
consistent, accurate, and useful for diverse applications. For example, 
well\mbox{-}curated data is essential for machine learning systems, which 
require high\mbox{-}quality data to produce reliable insights 
\citep{polyzotis2019data, jain2020overview}. A critical component of 
data curation is data cleaning, which involves identifying and rectifying 
inconsistencies, errors, and inaccuracies in datasets. While open data 
initiatives have made vast amounts of data available, ensuring the 
reliability of such data hinges on rigorous data cleaning processes. 
Efficient usage of software tools can significantly streamline this 
aspect of curation \citep[e.g.,][]{cody2017cody, van2018statistical}. Poor 
curation may result in issues such as missing data, formatting errors, 
or inconsistent values, leading to biased or inaccurate outcomes 
\citep{geiger2020garbage}. This is particularly critical in domains 
where machine learning is applied to sensitive tasks, such as public 
health or policy \citep{rahm2000data}.

Research into data curation has explored these challenges in\mbox{-}depth. 
Among the earliest discussions, \citet{witt2009constructing} focused 
on developing data curation profiles tailored to specific contexts, 
setting a precedent for targeted data management strategies. 
Addressing broader challenges in data sharing and management, 
\citet{borgman2012conundrum} highlighted the complexities of 
research data distribution, emphasizing the need for robust 
strategies. This is complemented by \citet{hart2016ten}, who outlined 
essential principles for effective data management, particularly 
emphasizing meticulous curation practices. The utility of curated 
open data is vividly illustrated in public health and global challenges, 
where \citet{cantor2018facets} demonstrated the utility of curated 
data in evaluating community health determinants, and 
\citet{shankar2021data} observed its critical role during the 
COVID\mbox{-}19 pandemic in managing collective responses.

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request (SR) Data as a case study. This dataset serves as a prime 
example for examining key issues in data curation, including data 
validity, consistency, and curation efficiency. We illustrate these 
points with live examples drawn from our 
processing of the 311 SR data. Secondly, building upon insights 
gained from this case study, we propose a set of data curation 
principles tailored for government\mbox{-}released open data. These 
principles are designed to address the unique challenges 
and requirements observed in the curation of such datasets.

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over a 2\mbox{-}year period from 2022 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
 also investigates field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines some concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}.

Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 SR Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non\mbox{-}emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non\mbox{-}emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone\mbox{-}based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID\mbox{-}19 pandemic, the 2021 expansion to include the MTA's city
subway system (the biggest expansion in their history), and a 
record 3.23 million service requests in 2023.

Today, the NYC 311 data system manages over 3 million service
requests (SRs) per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo\mbox{-}mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, provides technological support for the
open data system and data portal. Input data for the 311 system 
is sourced from 17 different NYC Agencies. Some of these Agencies
utilize the core 311 software, while others transfer data to the
311 Open Data Warehouse from in\mbox{-}house systems.

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping ity governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID\mbox{-}19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies

Our investigation utilizes an SR dataset covering  2\mbox{-}years (2022-2023).
This dataset is approximately 3.7 GB in size and contains over 
6.4 million rows with each row representing one SR. It contains 
41 columns of data and was collected on 15 September 2023. The 
dataset is used to conduct detailed analysis 
in areas such as data consistency, data validity, redundant fields, 
etc.. Instructions on downloading this file from the 
NYC Open Data Portal are available in the supplemental material.

During the 10\mbox{-}year period from 2014 to 2023,  311 experienced
a 50\% increase in the number of Service Requests (SRs), 
reflecting a growing public reliance and usage of 
the 311 system. (During this 10\mbox{-}year time-frame, the 
population of NYC grew just 0.6\%.). The most significant increase
in activity occurred in 2020, when the volume of requests 
surged to 3.23 million during the COVID\mbox{-}19 pandemic, 
illustrating the essential role of the 311 system in supporting 
NYC residents during times of crisis. This steady 
rise in SRs can be at least partially attributed to the increased 
accessibility of the 311 system via online and mobile 
platforms, as well as heightened public awareness of the service. 
While the spike in 2020 was exceptional, the broader trend 
indicates sustained growth in SR volume in future years,  
highlighting the system's expanding role in managing 
both routine City operations and extraordinary events.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
the responsible Agency, showing the cumulative 
percentage of SRs handled by each
Agency over a the 2\mbox{-}year time-frame. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

\begin{itemize}
    \item New York Police Department (NYPD) - 43\% of all SRs
    \item Housing Preservation and Development (HPD) - 21\%
    \item New York City Department of Sanitation (DSNY) - 10\%
    \item Department of Transportation (DOT) - 7\%
    \item Department of Environmental Protection (DEP) - 5\%
    \item Department of Parks and Recreation (DPR) - 4\%
\end{itemize}

Collectively these six City Agencies handle over 90\% of the total SR 
volume, indicative of the critical role these Agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The remaining
10\% of SRs is distributed across 10 additional Agencies. 
This concentration of SRs to the ``big six'' Agencies 
underscores the necessity for optimal data handling and 
curation processes within these six high volume Agencies.

@@

Figure~\ref{fig:SR_complaints}
illustrates how the top 20 complaint types account for 70\% 
of all SRs. Note that the distribution of complaints is skewed, 
with a small number of issues dominating total SR volume. 
Of special interest are noise\mbox{-}related complaints, of which there are eight different
types including residential, commercial, and 
street noise. When combined noise\mbox{-}related SRs make up 22\% of all 
complaints; the most frequent issue in the NYC 311 system. 
Other prominent categories include illegal parking, heat/hot water 
complaints, and blocked driveways. The cumulative percentage curve 
reveals that the top 20 complaint types comprise 70\% of SRs,
while the remaining 190 complaint types are spread thinly 
across the remaining 30\%. This suggests that improving 
responses to the most frequent complaints could have an 
outsized impact on overall service efficiency and resident 
satisfaction. This chart also provides insights into the 
operational pressures faced by City Agencies 
responsible for handling these high-volume complaints.

@@

\section{NYC 311 SR Data Cleansing Issues} 
\label{sec:issues}
Data cleansing refers to the process of identifying and rectifying
errors, inconsistencies, and inaccuracies within datasets to ensure
they are of high quality and reliable for analysis
\citep{maletic2005data, hosseinzadeh2023data}. The process
typically involves removing duplicate records, handling missing or
incomplete data, correcting mislabeled or inaccurate entries, and
standardizing data formats \citep[e.g.,][]{cody2017cody,
  van2018statistical}. In the context of open data, cleansing is
especially important as open datasets often come from diverse,
often uncoordinated sources, leading to variations in data quality,
completeness, and consistency. Without thorough data cleansing, 
the utility of open data can be limited and perhaps 
untrustworthy, affecting its reliability for research, 
policy making, and innovation. The main purpose of cleansing 
open data is to ensure that it is accurate, consistent, and usable a
cross multiple analytical platforms and by various stakeholders with a 
myriad of purposes. Data cleansing improves the trustworthiness 
of the data and enables more accurate analysis, 
better decision making, and the improved integration of 
data into machine learning models or other systems.

Many quality criteria are employed to ensure high\mbox{-}quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats; typically 
outlined in a \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}. An 
essential aspect of validation 
is domain compliance, where fields must adhere to a predefined set of 
values such as statuses, state names, or zip codes. Structural 
errors can also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary 
and the inconsistent use of blanks, spaces, NA, N/A, 
or ``<NA>'' to indicate missing data. Furthermore, redundant 
or irrelevant fields can clutter datasets, reducing efficiency and 
creating confusion or errors. Logical inconsistencies are another 
consideration, such as related fields that violate expected 
relationships, e.g. a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure meaningful data.

Here we identify the presence of issues in the NYC 311 SR open dataset;
not attempting to solve them, but rather to highlight their scope and
magnitude. We would recommend that a solution effort 
should be undertaken only after further investigation as to 
the why and how such issues came about, including a
discussion as to whether or not it is an actual \textit{error} or 
holds some other status. Coordination with the originating Agency
is, of course, paramount.

\subsection{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can make 
it difficult to analyze the data effectively. Here are some overall 
characteristics of the 311 SR data set:

\begin{itemize}
	\item There are 41 columns of data for each row, exportable as a CSV file.
	Each row represents a single Service Request (SR).

@@

	\item There are seven street fields; two pair of which appear to be duplicates

@@

	\item A free\mbox{-}form text field, \texttt{resolution\_description}, which 
	supports 934 characters of input, including commas and special characters.
\end{itemize}

The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' and include:

@@

While one can likely infer the meaning of these ``computed'' 
fields, uncertainty remains as to their derivation and 
usage. These six fields are not included in
the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}. Our 
analysis shows that these six ``computed'' fields have 
significant data validity issues, leading us to treat these six fields 
as experimental and not for suitable for normal use. Accordingly, 
we will not address these fields in our analysis.

\subsection{Missing Data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} fields are blank. When sorting fields by
blank or N/A values, they appear to divide into three groups:

@@

The Mostly Empty category includes such fields as
\texttt{taxi\_company\_borough}, \texttt{due\_date},
\texttt{road\_ramp}, and \texttt{bridge\_highway\_name}.
The Partially Empty category includes such fields as
\texttt{location\_type}, \texttt{landmark}, 
and \texttt{cross\_street\_1 \& \_2}. And the Few/None Empty category
 includes \texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. For some analysis efforts, it may be
prudent to inquire as to why some fields are mostly or almost always blank.
Figure~\ref{fig:blank_fields} presents a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

@@

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. The \texttt{latitude} 
and \texttt{longitude} fields were found to fall within the 
geographic boundaries of New York City, and 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} specifies very few domains of 
acceptable values for other fields. Nonetheless, the following fields were tested and found to comply with their expected domains, as determined by public 
usage and as evidenced in other historical datasets:

@@

However,  some fields proved problematic as to their 
compliance with a domain of allowable values. These include:

\paragraph{Zip Codes:}
\label{sec:zipcodesissues}
All \texttt{incident\_zip} zip codes should 
validate against the USPS database, which contains 
44,173 valid zip codes. And while the \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by Agency, it mirrors the \textit{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

\paragraph{Created and Closed Dates}
\label{sec:negativeduration}
With the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields would be automatically populated by the  
application software. Thus when updating an SR status to ``new'' or ``closed''
the appropriate software should automatically populate those fields. 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the \texttt{created\_date} and \texttt{closed\_date} fields that
an automated process would normally exclude, including:

\begin{itemize}
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date} resulting a nonsensical ``negative duration''.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past, e.g. ``1900-01-01''.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match \textbf{to the second} creating an impossible ``zero duration''.
    \item An unusually large number of SRs closed and/or created exactly at midnight 
    and noon, to the second.
\end{itemize}

\paragraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the \texttt{created\_date}(s) and 
\texttt{closed\_date}(s) to measure the \textit{duration} of SRs, 
a useful surrogate to measure an Agency's responsiveness. 
While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.  Duration is one of 
most frequently used 311metrics and is frequently analyzed, such as determining if 
one NYC borough receives faster response than another for certain 
complaints. There are 12,450 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Although this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses. Eight SRs with extremely 
large negative durations ($-$730 days), all originating from the 
Department of Homeless Services (DHS); all contain an entry 
of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Many
other SRs have negative durations in excess of 300+ days. Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis.

Excluding the extreme negative values ($-$730 days), 
Figure~\ref{fig:negative-duration-violin} shows the broad spread of 
negative duration SRs. While there are few outliers, the magnitude 
of these negative durations (300 days) is troubling and can produce 
bizarre and unwanted analytical results.
